---
title: "Akula-DATA607-Week01-Homework"
author: "Pavan Akula"
date: "February 6, 2017"
output: html_document
---

Loading the data about mushrooms from website https://archive.ics.uci.edu/ml/machine-learning-databases/mushroom/agaricus-lepiota.data

```{r get-data-from-url, echo=TRUE}
Mushrooms <- read.table("https://archive.ics.uci.edu/ml/machine-learning-databases/mushroom/agaricus-lepiota.data",header = FALSE,sep=",")
head(Mushrooms,10)
```

File can be loaded from local directory also.

```{r get-data-from-local-dir, echo=TRUE}
library(stringr)
localdir <- getwd()
destfile <- paste(str_trim(localdir),"//agaricus-lepiota.data",sep="")
Mushrooms <- read.table(destfile,header = FALSE,sep=",")
head(Mushrooms,10)
```


File contains 23 columns and 8124 rows. First column is categorical data showing mushroom catagory (e = edible and p = poisonous). Rest of 22 columns are attributes of mushrooms.

```{r info-data, echo=TRUE}
dim(Mushrooms)
str(Mushrooms)
```

Downloading data dictionary located at https://archive.ics.uci.edu/ml/machine-learning-databases/mushroom/agaricus-lepiota.names

```{r download-data-dic, echo=TRUE}
library(stringr)
localdir <- getwd()
destfile <- paste(str_trim(localdir),"//agaricus-lepiota.txt",sep="")
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/mushroom/agaricus-lepiota.names"
download.file(url, destfile, mode="wb")
```

Changing the column name to meaning full names. Example V1 = class, V2 = cap-shape and so on. These columns names are derived from data dictionary

```{r rename-cols, echo=TRUE}
names(Mushrooms)[names(Mushrooms) == "V1"] <- "class"
names(Mushrooms)[names(Mushrooms) == "V2"] <- "cap-shape"
```

```{r rename-cols-all, echo=TRUE}
library(plyr)
Mushrooms <- rename(Mushrooms,
c("V3"="cap-surface", 
"V4"="cap-color",  
"V5"="bruises",  
"V6"="odor",  
"V7"="gill-attachment",  
"V8"="gill-spacing",  
"V9"="gill-size",  
"V10"="gill-color",  
"V11"="stalk-shape",  
"V12"="stalk-root",  
"V13"="stalk-surface-above-ring",  
"V14"="stalk-surface-below-ring",  
"V15"="stalk-color-above-ring",  
"V16"="stalk-color-below-ring",  
"V17"="veil-type",  
"V18"="veil-color",  
"V19"="ring-number",  
"V20"="ring-type",  
"V21"="spore-print-color",  
"V22"="population",  
"V23"="habitat"
))
head(Mushrooms,10)
```

List distinct values in class column.

```{r unique-col-data1, echo=TRUE}
unique(Mushrooms$class)
```

Formatting data, updated coded values to actual values. Example: in class column e = edible and p = poisonous. Since there are only 2 levels data can be formated using "elseif""

```{r format-col-data1, echo=TRUE}
unique(Mushrooms$class)
Mushrooms$class <- ifelse(Mushrooms$class=="e","edible","poisonous")
unique(Mushrooms$class)
```

Column cap-shape contain more then two levels. Formating cap-shape data.

```{r format-col-data2, echo=TRUE}
unique(Mushrooms$`cap-shape`)
revalue(Mushrooms$`cap-shape`,c("b" = "bell")) -> Mushrooms$`cap-shape`
revalue(Mushrooms$`cap-shape`,c("c" = "conical")) -> Mushrooms$`cap-shape`
revalue(Mushrooms$`cap-shape`,c("x" = "convex")) -> Mushrooms$`cap-shape`
revalue(Mushrooms$`cap-shape`,c("f" = "flat")) -> Mushrooms$`cap-shape`
revalue(Mushrooms$`cap-shape`,c("k" = "knobbed")) -> Mushrooms$`cap-shape`
revalue(Mushrooms$`cap-shape`,c("s" = "sunken")) -> Mushrooms$`cap-shape`
unique(Mushrooms$`cap-shape`)
```

Formating cap-surface data.

```{r format-col-data3, echo=TRUE}
unique(Mushrooms$`cap-surface`)
revalue(Mushrooms$`cap-surface`,c("s" = "smooth")) -> Mushrooms$`cap-surface`
revalue(Mushrooms$`cap-surface`,c("y" = "scaly")) -> Mushrooms$`cap-surface`
revalue(Mushrooms$`cap-surface`,c("f" = "fibrous")) -> Mushrooms$`cap-surface`
revalue(Mushrooms$`cap-surface`,c("g" = "grooves")) -> Mushrooms$`cap-surface`
unique(Mushrooms$`cap-surface`)
```

Formating cap-color data.

```{r format-col-data4, echo=TRUE}
unique(Mushrooms$`cap-color`)
revalue(Mushrooms$`cap-color`,c("n" = "brown")) -> Mushrooms$`cap-color`
revalue(Mushrooms$`cap-color`,c("b" = "buff")) -> Mushrooms$`cap-color`
revalue(Mushrooms$`cap-color`,c("c" = "cinnamon")) -> Mushrooms$`cap-color`
revalue(Mushrooms$`cap-color`,c("g" = "gray")) -> Mushrooms$`cap-color`
revalue(Mushrooms$`cap-color`,c("r" = "green")) -> Mushrooms$`cap-color`
revalue(Mushrooms$`cap-color`,c("p" = "pink")) -> Mushrooms$`cap-color`
revalue(Mushrooms$`cap-color`,c("u" = "purple")) -> Mushrooms$`cap-color`
revalue(Mushrooms$`cap-color`,c("e" = "red")) -> Mushrooms$`cap-color`
revalue(Mushrooms$`cap-color`,c("w" = "white")) -> Mushrooms$`cap-color`
revalue(Mushrooms$`cap-color`,c("y" = "yellow")) -> Mushrooms$`cap-color`
unique(Mushrooms$`cap-color`)
```

Formating bruises data.

```{r format-col-data5, echo=TRUE}
unique(Mushrooms$`bruises`)
revalue(Mushrooms$`bruises`,c("t" = "bruises")) -> Mushrooms$`bruises`
revalue(Mushrooms$`bruises`,c("f" = "no")) -> Mushrooms$`bruises`
unique(Mushrooms$`bruises`)
```

Subset data where class = edible. Selecting first 5 columns

```{r subset-data1, echo=TRUE}
edible <- Mushrooms[Mushrooms$class == "edible",1:5]
head(edible,20)
```

Subset data where class = poisonous and bruises = no. Selecting columns based on names

```{r subset-data2, echo=TRUE}
poisonous <- subset(Mushrooms, Mushrooms$class == "poisonous" & Mushrooms$bruises == "no",select = c(class,`cap-color`,`cap-shape`,bruises))
head(poisonous,20)
```
